---
title: "Firearm Violence and Death Analysis"
author: "Anjali Chappidi"
date: '`r Sys.Date()`'
output: html_document
---
## Firearm Violence and Death Analysis {.tabset}

###Overview

The goal of this report is to perform an exploratory analysis on the Gun Deaths that happened in the year of 2013. 

The unstructured data used in this report is loaded from:
[Data Source](ftp://ftp.cdc.gov/pub/Health_Statistics/NCHS/Datasets/DVS/mortality). 
The Codebook for the above specified data can be found at:
[Code Book](http://www.cdc.gov/nchs/data/dvs/Record_Layout_2013.pdf)

Since the data loaded has various causes of deaths, it is subset to load FIREARM deaths data. Here we have subsetted various underlying causes of usage of Firearms such as :

1. Accidental  
2. Suicide
3. Homicide
4. Undetermined Intent and
5. Legal Intervention

I have attempted to slice the data as per my analysis requirements.

###Packages Required 

The file that is directly loaded without cleaning has 2601452 observations of 45 variables.
The description of each variable is provided in the codebook.

Some of the variables that I would like to describe are given below:

detail_age : GIves the age 
Underlying_Cause: Has some codes each of which describes a cause of death

The prerequisite libraries that are used are given as follows:

Description of packages:

1.DT = For HTML display   
2.readr = reads tabular data into R   
3.dplyr = data manipulation in r    
4.tidyr = Helps in tidying the data with magrittr    
5.ggplot2 = graphical representation in r    
6.tibble = For manipulating and printing dataframes    
7.magrittr = For using the pipe operator     
8.sqldf = For using sql commands    



```{r, message=FALSE, warning=FALSE}
library(DT)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
library(magrittr)
library(sqldf)

```

###Data Description and Import

The data is first downloaded from the source and structured

```{r, message=FALSE, warning=FALSE}
year <- 2013
url <- "ftp://ftp.cdc.gov/pub/Health_Statistics/NCHS/Datasets/DVS/mortality/mort2013us.zip"

structure1 <- fwf_widths(c(19,1,40,2,1,1,2,2,1,4,1,2,2,2,2,1,1,1,16,4,1,1,1,1,34,1,1,4,3,1,3,3,2,1,281,1,2,1,1,1,1,33,3,1,1),
                       col_names = c("drop1", "res_status", "drop2", "education_89", "education_03", "education_flag", "month", 
                                     "drop3", "sex", "detail_age", "age_flag", "age_recode", "age_recode2", "age_group", 
                                     "age_infant", "death_place", "marital", "day_of_week", "drop4", "data_year", "at_work", 
                                     "death_manner", "burial", "autopsy", "drop5", "activity", "injury_place", 
                                     "underlying_cause", "cause_recode358", "drop6", "cause_recode113", "cause_recode130", 
                                     "cause_recode39", "drop7", "multiple_causes", "drop8", "race", "race_bridged", "race_flag", 
                                     "race_recode", "race_recode2", "drop9", "hispanic", "drop10", "hispanic_recode"))
  
  temp <- tempfile()
  download.file(url, temp, quiet = T)
```

After the initial download , it is unzipped and structured. Below is a overview at the data.

```{r, message=FALSE, warning=FALSE} 

  # Read in data
  Rawfire1 <- read_fwf(unzip(temp), structure1)
  head(Rawfire1)


```

A look at the initial data:

```{r, message=FALSE, warning=FALSE}

DT::datatable(head(Rawfire1,100))

```

Conert the data into tibble format:

```{r, message=FALSE, warning=FALSE}
Raw_Tibble <- as_tibble(Rawfire1)
dim(Raw_Tibble)
```

###Data Subsetting and Cleaning

Looking at the data, we need to perform certain actions on it:

1. Dropping empty fields. Now the number of variables drop to 35. 

```{r, message=FALSE, warning=FALSE}
Raw_Tibble <- Raw_Tibble %>% select(-contains("drop"))
```

Subsetting Raw data file based on the underlying cause of death having codes of Firearms.
1. Accidental (W32,W33,W34)
2. Suicide (X72,X73,X74)
3. Homicide (U014,X93,X94)
4. Undetermined Intent (Y22,Y23,Y24)and
5. Legal Intervention (Y350)

  
```{r, message=FALSE, warning=FALSE}
Firearm <- Raw_Tibble %>% filter(underlying_cause %in% c("W32", "W33", "W34", "X72", "X73", "X74", "U014", "X93", "X94", "X95", "Y22", "Y23", "Y24", "Y350"))

head(Firearm)
```

Now I would like to categorise them and mutate the data based on the underlying cause by adding the categorical variables for intent, weapon, plus dummy for police shootings.

```{r, message=FALSE, warning=FALSE}
Firearms <- Firearm %>%
    mutate(IntentDeath = ifelse(underlying_cause %in% c("W32", "W33", "W34"), "Accidental",
                    ifelse(underlying_cause %in% c("X72", "X73", "X74"), "Suicide",
                    ifelse(underlying_cause %in% c("*U01.4", "X93", "X94","X95","Y350"),"Homicide",
                    ifelse(underlying_cause %in% c("Y22", "Y23", "Y24"), "Undetermined", 
                           ifelse(underlying_cause == "Y350", "PoliceDeath",NA))))),
           WeaponDeaths = ifelse(underlying_cause %in% c("W32", "X72", "X93", "Y22"), "Handgun",
              ifelse(underlying_cause %in% c("W33", "X73", "X94", "Y23"), "Rifle","Other/unknown")),
           Year = year)
head(Firearms)
```

Next step of cleaning the data is Creating a new age variable by coding every age under 1 as "0".
A look at the data after this step of cleaning:

```{r, message=FALSE, warning=FALSE}

  Firearms <- Firearms %>%
    mutate(age = ifelse(substr(detail_age, 1, 1) == "1", as.numeric(substr(detail_age, 2, 4)), # Year
                        ifelse(detail_age == 9999, NA, 0)),
           age = ifelse(age == 999, NA, age))

DT::datatable(head(Firearms,100))

```

###Exploratory Data Analysis:


Below is a step to subset the data as per the requirement for analysis:

From the codebbok , the following analysis of the variables are drawn and hence mutated with the new columns for easy understanding and analysis.

The new columns added for the analysis are:
Place, education, race.

```{r, message=FALSE, warning=FALSE}

Firearms <- Firearms %>% 
  mutate(place = factor(injury_place, labels = c("Home", "Residential institution", "School/instiution", "Sports", "Street","Trade/service area", "Industrial/construction", "Farm", "Other specified","Other unspecified")),
         education = ifelse(education_flag == 1, 
                            cut(as.numeric(education_03), breaks = c(0, 2, 3, 5, 8, 9, labels = c("Less than HS", "HS/GED", "Some college", "BA+", NA))),
                            cut(as.numeric(education_89), breaks = c(0, 11, 12, 15, 17, 99), labels = c("Less than HS", "HS/GED", "Some college", "BA+", NA))),
         race = ifelse(hispanic > 199 & hispanic <996, "Hispanic",
                       ifelse(race == "01", "White",
                              ifelse(race == "02", "Black",
                                     ifelse(as.numeric(race) >= 4 & as.numeric(race) <= 78, "Asian/Pacific Islander","Native American/Alaskan")))),
         race = ifelse(is.na(race), "Unknown", race))
Firearms1 <- sqldf("select year, month, IntentDeath,WeaponDeaths,sex, age, race,place, education from Firearms")
head(Firearms1)

```

The Main aim of the analysis is to explore on:
1. The underlying causes behind the deaths that occurred and what mainly is the cause behind the deaths
2. How the deaths differ between different sex
3. How the deaths differ in different age groups
4. How the deaths differ in different races


```{r, message=FALSE, warning=FALSE}

Firearms1 %>% ggplot(aes(x=age))+geom_histogram(aes(y = ..density..),binwidth = 10, color ="grey30", fill= "lightcyan") + geom_density(alpha = .2, fill = "khaki3")+ xlab("Age")+ ylab("Count of Deaths")


F1 <- Firearms1 %>% group_by(place) %>% summarize(n = n()) %>% mutate(frequency = n/ sum(n) ) %>% arrange(desc(n))
F1 %>% ggplot(aes(x=place,y=frequency, fill=n))+theme(axis.text.x = element_text(angle = 45, size=8, hjust = 1, vjust = 1))+geom_bar(stat = "identity")

F2 <- Firearms1 %>% group_by(education,sex) %>% summarize(n = n()) %>% mutate(frequency = n/ sum(n) ) %>% arrange(desc(n))
F2 %>% ggplot(aes(x=education,y=frequency, fill=n))+geom_bar(stat = "identity")

F3 <- Firearms1 %>% group_by(education,sex) %>% summarize(n = n()) 
F3 %>% ggplot(aes(x=sex,y=n)) + geom_bar(stat = "identity", color ="grey30",fill = "lightcyan") + facet_wrap(~education,scales = "free_y")

F4 <- Firearms1 %>% group_by(race,sex) %>% summarize(n = n()) 
F4 %>% ggplot(aes(x=sex,y=n)) + geom_bar(stat = "identity", color ="grey30",fill = "lightcyan") + facet_wrap(~race,scales = "free_y")

F5 <- Firearms1 %>% group_by(IntentDeath,sex) %>% summarize(n = n()) %>% mutate(frequency = n/ sum(n) ) %>% arrange(desc(n))
F5 %>% ggplot(aes(x = sex, y = n, fill = sex)) + geom_boxplot()
F5 %>% ggplot(aes(x=sex,y=n)) + geom_bar(stat = "identity", color ="grey30",fill = "lightcyan") + facet_wrap(~IntentDeath,scales = "free_y")

F6 <- Firearms1 %>% group_by(IntentDeath,race) %>% summarize(n = n())
F6 %>% ggplot(aes(x=race,y=n)) +theme(axis.text.x = element_text(angle = 45, size=8, hjust = 1, vjust = 1))+ geom_bar(stat = "identity", color ="grey30",fill = "lightcyan") + facet_wrap(~IntentDeath,scales = "free_y")
```

###Summary

Observations:

1.Women are far less likely to be victims in any race.     
2.It is observed that nearly 2/3rds of the total deaths are suicides.   
3.More than 85% of suicide victims are males.   
4.More than half of the suicides are men with more age.   
5.White men have the highest number of deaths.    
